---
title: "Project 1"
author: "Cynthia Kulak cek2338"
date: "3/15/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#Data sets
BodyMass<-read.csv("BodyMass.csv")
Habitat<-read.csv("Habitat.csv")
```

*Introduction: The datasets I have chosen show the body mass and habitats of the world's primates from a review of 1,216 published studies from 1941-2018. I have always been interested in primates and I am considering becoming a primatologist, so I thought this information would be interesting. Both datasets come from zenodo.org, which I found from the shared google sheet of datasets. Both datasets share the variables Family, Genus, CommonName, Species, and Species (ITIS), and the references used for the data. Additional variables in the Body Mass dataset are the average body mass, the average body mass of males and females (all in kilograms). Additional variables in the Habitat dataset are forest, savanna, shrubland, grassland, wetlands, rocky areas, and desert. Once these datasets are combined and analyzed, I think there will be associations between each habitat and the average body mass.*


```{R}
#Tidying
library(dplyr)
library(tidyverse)
habitat_tidy<-Habitat%>%pivot_longer(c(6:12), names_to="habitat", values_to = "value")%>%
  mutate(habitat = recode(habitat, "Habitat_Forest"="forest",
                          "Habitat_Savanna"="savanna",
                          "Habitat_Shrubland"="shrubland",
                          "Habitat_Grassland"="grassland",
                          "Habitat_Wetlands"="wetlands",
                          "Habitat_Rocky.areas"="rocky.areas",
                          "Habitat_Desert"="desert"))%>%
  filter(value=="1")%>%select(-value)
glimpse(habitat_tidy)
habitat_tidy_1<-Habitat%>%pivot_longer(c(6:12), names_to="habitat", values_to = "value")%>%
  mutate(habitat = recode(habitat, "Habitat_Forest"="forest",
                          "Habitat_Savanna"="savanna",
                          "Habitat_Shrubland"="shrubland",
                          "Habitat_Grassland"="grassland",
                          "Habitat_Wetlands"="wetlands",
                          "Habitat_Rocky.areas"="rocky.areas",
                          "Habitat_Desert"="desert"))
glimpse(habitat_tidy_1)
habitat_1<-habitat_tidy_1%>%pivot_wider(names_from="habitat", values_from = "value")
summary(habitat_1)
```

*Both datasets were already tidy, but I wanted to see what would happen if I changed up the Habitat dataset by combining all the habitats into one column, with the name of the habitat as the values. I achieved this by using pivot_longer to make "habitat" as a new column with "habitat_(insert habitat here) as the values. To get rid of "habitat"" in the value names, I used mutate with recode so just the type of habitat was left. Then I used filter to get rid of any rows that weren't in that habitat (have a 0 as the value) and then used select to get rid of the value column altogether. I realized this wasn't as good as the original dataset, so I ran habitat_tidy again, but this time without getting filtering the value column or geeting rid of the value column. Then I used pivot_wider to make each habitat it's own column with numerical values to designate which habitats the primates reside.*


```{R}
#Joining
mass_and_habitat<-inner_join(habitat_1, BodyMass, by = c("Family", "Genus", "CommonName", "Species", "Species..ITIS."))
summary(mass_and_habitat)
```

*I decided to join the datasets by the common variables of Family, Genus, CommonName, Species, and Species (ITIS). Through this joining, I can easily compare habitats, body mass, and refs.*


```{r}
#Wrangling
#mutate
ratio<-mass_and_habitat%>%mutate(BodyMassRatio = BodyMassMale_kg/BodyMassFemale_kg)
summary(ratio)
#Filter
Tenkg<-ratio%>%filter(BodyMass_kg>10)
summary(Tenkg)
#Select
Tenkg_Ratio<-Tenkg%>%select(3, 8:17, 19)
summary(Tenkg_Ratio)
#arrange
arrange_m_and_h<-ratio%>%arrange(desc(BodyMass_kg))
summary(arrange_m_and_h)
#group_by, summarize, and summary statistics
  #summary statistics of average body mass for each Common Name
summ_stats_BodyMass<-ratio%>%
  group_by(Species..ITIS.)%>%
  summarize(mean_BodyMass=mean(BodyMass_kg,na.rm=T), 
            sd_BodyMass=sd(BodyMass_kg, na.rm=T), 
            var_BodyMass=var(BodyMass_kg, na.rm =T),
            med_BodyMass=median(BodyMass_kg, na.rm=T))
summary(summ_stats_BodyMass)
  #summary statistics of average male body mass for each Common Name
summ_stats_MaleBodyMass<-ratio%>%
  group_by(Species..ITIS.)%>%
  summarize(mean_MaleBodyMass=mean(BodyMassMale_kg,na.rm=T), 
            sd_MaleBodyMass=sd(BodyMassMale_kg, na.rm=T), 
            var_MaleBodyMass=var(BodyMassMale_kg, na.rm =T),
            med_MaleBodyMass=median(BodyMassMale_kg, na.rm=T))
summary(summ_stats_MaleBodyMass)
  #summary statistics of average female body mass for each Common Name
summ_stats_FemaleBodyMass<-ratio%>%
  group_by(Species..ITIS.)%>%
  summarize(mean_FemaleBodyMass=mean(BodyMassFemale_kg,na.rm=T), 
            sd_FemaleBodyMass=sd(BodyMassFemale_kg, na.rm=T), 
            var_FemaleBodyMass=var(BodyMassFemale_kg, na.rm =T),
            med_FemaleBodyMass=median(BodyMassFemale_kg, na.rm=T))
summary(summ_stats_FemaleBodyMass)
  #mean body mass of each family in wetlands habitat
mean_WetlandsBodyMass<-ratio%>%
  group_by(Family, wetlands)%>%
  summarize(mean_BodyMass=mean(BodyMass_kg, na.rm=T))
summary(mean_WetlandsBodyMass)
```

*For the mutate function, I divided male body mass by female body mass to create a new column that gives the body mass ratio for the primates. Then I used filter to see how many primates had an average body mass over 10kg. Then I wanted to see directly what habitats and what the body mass ratio was for the primates with an average body mass over 10kg, so I used the select function for just the common name, the habitats, and the body mass ratio. Then I went back to the dataset ratio to arrange the average body mass from greatest to least. I used group_by and summarize to get some of the summary statistics for all of the Body Mass variables for each Species...ITIS. I tried to include min and max in the summary statistics, but I got too many error warning so I left it out. I also grouped the categorical variables Family and Wetlands to get the mean Body Mass for each familt at that habitat. *


```{r}
#Visualizing
  #correlation heatmap
    #compute correlation matrix of body masses
corrmat<-mass_and_habitat%>%na.omit%>%select(8:17)
mat<-cor(corrmat)
    #tidy
tidycor<-mat%>%as.data.frame%>%
  rownames_to_column%>%
  pivot_longer(-1,names_to="name",values_to="correlation")
    #graph
tidycor%>%ggplot(aes(rowname,name,fill=correlation))+
  geom_tile()+
  scale_fill_gradient2(low="red",mid="white",high="blue")+ 
  geom_text(aes(label=round(correlation,2)),color = "black", size = 4)+ 
  theme(axis.text.x = element_text(angle = 90, hjust=1))+ 
  coord_fixed()
  #plot 2
ggplot(data = mass_and_habitat, aes(x = Family, y = BodyMass_kg, color=Genus))+
  geom_point(size=3)
  #plot3
ggplot(ratio, aes(x = Family, y = BodyMassRatio))+
geom_bar(stat="summary",fun.y="mean", position="dodge")+
geom_errorbar(stat="summary",position="dodge")
```

*For the correlation heatmap, I looked at the correlation between body masses and habitats. I already know it's incorrect since there isn't data available for the correlation between body masses and 4 of the habitats, I'm just uncertain on how to fix it. But from this map, it looks like there is little correlation between body masses and the wetlands habitat, but moderate correlation between body masses and the savanna and shrubland habitats. When looking at the correlation between body mass, male body mass, and female body mass, it is clear that all of those have a pretty strong correlation.*

*The second plot is a simple scatter plot showing the Body Masses from each individual per family. From this graph, it is clear to see that the family Hominidae has a significantly larger body mass than any other family. Hominidae, Cercopithecidae, and Cebidae seem to have the greatest variance in body mass.I also added the Genus as the color of the dots, but because of their quanity and proximity to each other, it is difficult to distinguish anything other than which Genus has the greatest body mass in each family.*

*For plot 3 I graphed the mean body mass ratio for each family. This shows that for most of the families, the body mass ratio for males to females is 1, an only one family has a body mass ratio close to 2, which is tha same family with the largest overall body mass.*


```{r}
#Dimentionality Reduction
library(cluster)
library(GGally)
  #average silhouette width 
pam_dat<-mass_and_habitat%>%select(15:17)%>%na.omit
sil_width<-vector()
for(i in 2:10){
  pam_fit <- pam(pam_dat, k = i)
  sil_width[i] <- pam_fit$silinfo$avg.width
}
 ggplot()+geom_line(aes(x=1:10,y=sil_width))+scale_x_continuous(name="k",breaks=1:10)
pam1<-pam_dat %>% pam(k=3)
pam1
plot(pam1, which=2)
#I can't get this plot to work so I'm leaving it as a comment to show that I tried:(
  #mass_and_habitat%>% mutate(cluster=as.factor(pam1$clustering)) %>% ggpairs(columns = 15:17, aes(color=cluster))
```

*I ran k-means/PAM clustering and I found that 3 clusters should be chosen to maximize the silhouette width. I was not able to get the ggplot to work, but I found that the average silouette width is 0.62, which means, based on the rule-of-thumb thresholds, the final fit of this cluster is reasonable.*


```{R}